Sparse spectral methods for partial differential equations on spherical caps, highlights

### 2020-12-22 papers/spherical-caps/highlights.tex
Add statement on significance and highlights

\title{Sparse spectral methods for partial differential equations on spherical caps}
\author{Ben Snowball, Sheehan Olver\thanks{Department of Mathematics, Imperial College, London, England}}
%\date{}							% Activate to display a given date or no date

@@

\begin{abstract}
In recent years, sparse spectral methods for solving partial differential equations have been derived using hierarchies of classical orthogonal polynomials on intervals, disks, disk-slices and triangles. In this work we extend the methodology to a hierarchy of non-classical multivariate orthogonal polynomials on spherical caps. The entries of discretisations of partial differential operators can be effectively computed using formulae in terms of (non-classical) univariate orthogonal polynomials. We demonstrate the results on partial differential equations involving the spherical Laplacian and biharmonic operators, showing spectral convergence. 
\end{abstract}

%
\section{Introduction}

This paper develops sparse spectral methods for solving linear partial differential equations on certain subsets of the sphere---specifically spherical caps. More precisely, we consider the solution of partial differential equations on the spherical cap $\Omega$ defined by
\begin{align*}
	\Omega := \{(x,y,z) \in \R^3 \quad | \quad \alpha < z < \beta, \: x^2 + y^2 + z^2 = 1\}
\end{align*}
where $\alpha \in (-1,1)$ and $\beta := 1$.

\begin{remark}
For simplicity we focus on the case of a spherical cap, though there is an extension to a spherical band by taking $\beta \in (\alpha,1)$. The methods presented here translate to the spherical band case by including the necessary adjustments to the weights and recurrence relations we present in this paper. These adjustments make the mathematics more involved, which is why they are omitted here, but the approach is the same.
\end{remark}

We advocate using a basis that is polynomial in cartesian coordinates, that is, polynomial in $x$, $y$, and $z$, and orthogonal with respect to a prescribed weight: that is, multivariate orthogonal polynomials, whose construction was considered in \cite{olver2020orthogonal}. Equivalently, we can think of these as polynomials modulo the vanishing ideal $\{ x^2 + y^2 + z^2 = 1 \}$, or simply as a linear recombination of spherical harmonics that are orthogonalised on a subset of the sphere. This is in contrast to more standard approaches based on mapping the geometry to a simpler one (e.g., a rectangle or disk) and using orthogonal polynomials in the mapped coordinates (e.g., a basis that is polynomial in the spherical coordinates $\varphi$ and $\theta$). The benefit of the new approach is that we do not need to resolve Jacobians, and thereby we can achieve sparse discretisations for partial differential operators, including those with polynomial variable coefficients. Further, we avoid the singular nature at the poles or  as $\alpha$ approaches $0$ that  such a projection may give, since our new approach yields a smooth polynomial basis for all $\alpha \in [-1,1)$.

On the spherical cap, the family of weights we consider are of the form
\begin{align*}
	\Wa(x,y,z) := (z - \alpha)^a, \qfor (x,y,z) \in \Omega,
\end{align*}
noting that $\Wa(x,y,z) = 0$ for $(x,y,z) \in \partial \Omega$ when $a > 0$. The corresponding OPs denoted $\scopnkia(x,y,z)$, where $n$ denotes the polynomial degree, $0 \le k \le n$ and $i \in \{0, \min(1,k)\}$. We define these to be orthogonalised lexicographically, that is,
\begin{align*}
	\scopnkia(x,y,z) = C_{n,k,i} \: x^{k-i} \: y^i \: z^{n-k} + (\hbox{lower order terms})
\end{align*}
where $C_{n,k,i} \neq 0$ and \enquote{lower order terms} includes degree $n$ polynomials of the form $x^{j - i} \: y^{i} \: z^{n-j}$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs in Definition~\ref{def:OPconstruction}.

We consider partial differential operators involving the spherical Laplacian (the Laplace--Beltrami operator): in spherical coordinates 
\begin{align*}
	z &= \cos\varphi, \\
	x &= \sin\varphi \cos\theta = \rho(z) \cos\theta, \\
	y &= \sin\varphi \sin\theta = \rho(z) \sin\theta.
\end{align*}
where $ \rho(z) := \sqrt{1-z^2}$, we have
\begin{align*}
	\DeltaS &= {1 \over \sinphi} \ppphi \Big( \sinphi \ppphi \Big) + {1 \over \sin^2 \varphi} \ppthetatwo = {1 \over \rho} \ppphi \Big( \rho \ppphi \Big) + {1 \over \rho^2} \ppthetatwo
\end{align*}
i.e. $\DeltaS f(\xvec) = \Delta f({\xvec \over \norm{\xvec}})$ for $\xvec := (x,y,z) \in \R^3$. We do so by considering the component operators $\rho {\partial \over \partial \varphi}$ and ${\partial \over \partial \theta}$ applied to OPs with a specific choices of weight so that their discretisation is sparse, see Theorem~\ref{theorem:sparsityofdifferentialoperators}.  Sparsity comes from expanding the domain and range of an operator using different choices of the parameter $a$, a la the ultraspherical spectral method for intervals \cite{olver2013fast}, triangles \cite{olver2019triangle} and disk-slices and trapeziums \cite{snowball2019sparse}, and the related work on sparse discretisations on disks \cite{vasil2016tensor} and spheres \cite{vasil2019tensor,lecoanet2019tensor}.  As in the  disk-slice case in 2D \cite{snowball2019sparse}, we use an integration-by-parts argument to deduce the sparsity structure.

 The three-dimensional orthogonal polynomials defined here involve the same non-classical (in fact, semi-classical \cite[\S5]{magnus1995painleve}) 1D OPs as those outlined for the disk-slice, and so methods for calculating these 1D OP recurrence coefficients and integrals have already been outlined \cite{snowball2019sparse}. In particular, by exploiting the connection with these 1D OPs we can construct discretizations of general partial differential operators of size $(p+1)^2 \times (p+1)^2$ in $O(p^3)$ operations, where $p$ is the total polynomial degree. This clearly compares favourably to proceeding in a na\"ive approach where one would require $O(p^6)$ operations.

Note that we consider partial differential operators that are not necessarily rotational invariant:  for example, one can use these techniques for  Schr\"odinger operators $\DeltaS + v(x,y,z)$ where $v$ is first approximated by a polynomial. A nice feature though is that if the partial differential operator is invariant with respect to rotation around the $z$ axis (e.g., a Schr\"odinger operator with potential $v(z)$) the discretisation decouples, and can be reordered as a block-diagonal matrix. This improves the complexity further to an optimal $O(p^2$), which is demonstrated in Figure \ref{fig:complexity} with $v(x,y,z) = \cos z$.

An overview of the paper is as follows:

\noindent \secref{OPs}: We present our definition of a (one-parameter) family of 3D orthogonal polynomials (OPs) on the spherical cap domain $\Omega$, by combining 1D OPs on the interval $(\alpha, 1)$ with Chebyshev polynomials, to form 3D OPs on the spherical cap surface. We show that these families will lead to sparse Jacobi operators for multiplication by $x, y, z$ and demonstrate how to obtain the 3D OPs.

\noindent\secref{PDOs}: We define several partial differential operators such as spherical Laplacians and  show that these will  be sparse when applied to a suitable choice of expansions in bases built from OPs on the spherical cap. We can exactly calculate the non-zero entries of these sparse operators using the quadrature rule associated with the non-classical 1D OPs.

\noindent\secref{Computation}: We derive a quadrature rule on the spherical cap that can be used to expand a function in the OP basis up to a given order $N$, and demonstrate how to evaluate a function using the Clenshaw algorithm using the coefficients of its expansion.

\noindent\secref{Examples}: We demonstrate the proposed technique for solving differential equations on the spherical cap such as the Poisson equation, variable coefficient Helmholtz equation, and Biharmonic equation.

%\noindent\appref{PFEM}: We outline how our framework can allow us to construct sparse $hp$-finite element method on the sphere, using spherical band and cap elements that would capture spherical geometry precisely. We lay the groundwork by simply presenting a sparse $p$-finite element method using a single spherical cap element.

\noindent{\bf Acknowledgments}: The first author was supported by the Engineering and Physical Sciences Research Council Mathematics of Planet Earth Centre for Doctoral Training at Imperial College London and the University of Reading, with grant number EP/L016613/1.The second author was supported by the Leverhulme Trust Research Project Grant RPG-2019-144 ``Constructive approximation theory on and inside algebraic curves and surfaces''.

\section{Orthogonal polynomials on spherical caps}\label{Section:OPs}

In this section we outline the construction and some basic properties of $\scopnkia(x,y,z)$.

\subsection{Explicit construction}

We can construct the 3D orthogonal polynomials on $\Omega$ from 1D orthogonal polynomials on the interval \([\alpha,\beta]\), and from Chebyshev polynomials. We do so in terms of Fourier series, which, following \cite{olver2020orthogonal}, we write here as orthogonal polynomials in $x$ and $y$:

\begin{definition}\label{def:Ydefinition}
	Define the unit circle $\omega := \{ \mathbf{x} = (x,y) \in \R^2 \: | \: x^2 + y^2 = 1\}$, and define the parameter $\theta$ for each $(x,y) \in \omega$ by $x = \cos\theta$, $y = \sin\theta$. Define the polynomials $\{\chki\}$ for $k = 0, 1, \dots$, $i = 0, 1$ on $(x,y) \in \omega$ by
\begin{align*}
	\ch_{0,0}(\xvec) \equiv \ch_{0,0}(x, y) &:= \ch_{0} =: \ch_{0,0}(\theta) \\
	\ch_{k,0}(\xvec) \equiv \ch_{k,0}(x, y) \ &:= T_k(x) = \cos k \theta =: \ch_{k,0}(\theta), \quad k = 1,2,3,\dots \\
	\ch_{k,1}(\xvec) \equiv \ch_{k,1}(x, y) &:= y \: U_{k-1}(x) = \sin k \theta =: \ch_{k,1}(\theta), \quad k = 1,2,3,\dots
\end{align*}
where $\ch_{0} := \frac{\sqrt{2}}{2}$ and $T_k$, $U_{k-1}$ are the standard Chebyshev polynomials on the interval $[-1,1]$. The $\{Y_{k,i}\}$ are orthonormal with respect to the inner product
\begin{align*}
	\ip< p, \: q >_{\ch} &:= \frac{1}{\pi} \: \int_0^{2\pi} p(\xvec(\theta)) \: q(\xvec(\theta)) \: \D \theta
\end{align*}
\end{definition}
Note that we have defined $\ch_0$ so as to ensure orthonormality.

\begin{proposition}[\cite{olver2020orthogonal}]\label{prop:construction}
	Let $w : (\alpha,\beta) \: \to \R$ be a weight function. For $n = 0,1,2,\dots, $ let $\{r_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k} w(x)$ where $0 \le k \le n$. Then the 3D polynomials defined on $\Omega$
\begin{align*}
	\scopnki(x,y,z) := r_{n-k,k}(z) \: \rho(z)^k \: \ch_{k,i}\fpr({x \over \rho(z)}, {y \over \rho(z)})
\end{align*}
for $i \in {0,1}, \: 0 \le k \le n, \: n = 0,1,2,\dots$ are orthogonal polynomials with respect to the inner product
\begin{align*}
	\ip< p, \: q > &:= \int_\Omega p(x,y,z) \: q(x,y,z) \: w(z) \: \D A \\
	&= \int_0^{\cos^{-1}(\alpha)} \int_0^{2\pi} p\big(\sinphi \costheta, \sinphi \sintheta, \cosphi \big) \: q\big(\sinphi \costheta, \sinphi \sintheta, \cosphi \big) \: \times \\
	&\qquad w(\cosphi) \: \sinphi \: \D \theta \: \D \varphi \\
	&= \int_\alpha^1 \int_0^{2\pi} p\big(\rho(z) \cos\theta, \rho(z) \sin\theta, z\big) \: q\big(\rho(z) \cos\theta, \rho(z) \sin \theta, z\big) \: w(z) \: \D \theta \: \D z
\end{align*}
on $\Omega$, where $\D A = \sinphi \: \D \theta \: \D \varphi$ is the uniform spherical measure on $\Omega$. 
\end{proposition}

For the spherical cap, we can use Proposition \ref{prop:construction} to create our one-parameter family of OPs. We first introduce notation for our family of non-classical univariate OPs that will be used as the $r_n$ polynomials above.
\begin{definition}[\cite{snowball2019sparse}]\label{def:OPconstruction}
	Let $\genjacw^{(a,b)}(x)$ be a weight function on the interval $(\alpha, 1)$ given by:
\begin{align*}
	\genjacw^{(a,b)}(x) &:= (x - \alpha)^{a} \: \rho(x)^{b}
\end{align*}
and define the associated inner product by:
\begin{align}
	\ip< p, \: q >_{\genjacw^{(a,b)}} &:= \frac{1}{\normgenjac^{(a,b)}} \: \int_\alpha^1 p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x \label{eqn:ipgenjac}
\end{align}
where
\begin{align}
	\normgenjac^{(a,b)} := \int_\alpha^1 \: \genjacw^{(a,b)}(x) \: \D x \label{eqn:ipnormalisation}
\end{align}
is a normalising constant.
Denote the two-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}).
\end{definition}

We can now define the 3D OPs for the spherical cap.
\begin{definition}\label{def:constuction}
	Define the one-parameter 3D orthogonal polynomials via:
\begin{align}
	\scopnkia(x,y,z) := \genjacnmk^{(a,2k)}(z) \: \rho(z)^k \: \chki\fpr(\frac{x}{\rho(z)}, \frac{y}{\rho(z)}), \quad (x,y,z) \in \Omega.
\end{align}
\end{definition}

By construction, $\{\scopnkia\}$ are orthogonal with respect to the inner product
\begin{align*}
	\ip< p, \: q >_{\scopa} &:= \int_\Omega p(\xvec, z) \: q(\xvec, z) \: \genjacw^{(a,0)}(z) \: \D A \\
	&= \int_\alpha^1 \int_0^{2\pi} p(\rho(z) \cos\theta, \rho(z) \sin\theta, z) \: q(\rho(z) \cos\theta, \rho(z) \sin\theta, z) \: \D \theta \: \genjacw^{(a,0)}(z) \: \D z,
\end{align*}
with
\begin{align}
	\norm{\scopnkia}_{\scopa}^2 &:= \ip< \scopnkia, \: \scopnkia >_{\scopa} = \pi \: \normgenjac^{(a, 2k)}. \label{eqn:normscop}
\end{align}
We note that the weight $\genjacw^{(a,b)}(z)$ has been used in the construction of 2D orthogonal polynomials on disk-slices and trapeziums \cite{snowball2019sparse}, where a method for obtaining recurrence coefficients and evaluating integrals was established (the weight is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}).

\subsection{Jacobi matrices}\label{subsection:jacobimats}

We can express the three-term recurrences associated with $\genjac_n^{(a,b)}$ as
\begin{align}
	x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x) \label{eqn:Rrecurrence}
\end{align}
where the coefficients are calculatable (see \cite{snowball2019sparse}). We can use (\ref{eqn:Rrecurrence}) to determine the 3D recurrences for $\scopnkia(x,y,z)$. Importantly, we can deduce sparsity in the recurrence relationships.  We first require the following lemma.

\begin{lemma}\label{lemma:Yrecurrence} 
The following identities hold for $k = 2,3,\dots$, $j = 0,1,\dots$ and $i, h \in \{0,1\}$:
\begin{align*}
	&1) \quad \int_0^{2\pi} \ch_0 \: \chjh(\theta) \: \cos\theta \: \D \theta = \ch_0 \: \pi \: \delta_{0,h} \: \delta_{1,j} \\
	&2) \quad \int_0^{2\pi} \ch_0 \: \chjh(\theta) \: \sin\theta \: \D \theta = \ch_0 \: \pi \: \delta_{1,h} \: \delta_{1,j} \\
	&3) \quad \int_0^{2\pi} \ch_{1,i}(\theta) \: \chjh(\theta) \: \cos\theta \: \D \theta = \pi \: \delta_{i,h} \: (\ch_0 \: \delta_{0,j} + \half \delta_{2,j}) \\
	&4) \quad \int_0^{2\pi} \ch_{1,i}(\theta) \: \chjh(\theta) \: \sin\theta \: \D \theta = \pi \:  \delta_{|i-1|,h} \: ((-1)^{i+1} \: \ch_0 \: \delta_{0,j} + (-1)^i \: \half \: \delta_{2,j}) \\
	&5) \quad \int_0^{2\pi} \chki(\theta) \: \chjh(\theta) \: \cos\theta \: \D \theta = \half \: \pi \: \delta_{i,h} \: (\delta_{k-1,j} + \delta_{k+1,j}) \\
	&6) \quad \int_0^{2\pi} \chki(\theta) \: \chjh(\theta) \: \sin\theta \: \D \theta = \half \: \pi \: \delta_{|i-1|,h} \: ((-1)^{i+1} \: \delta_{k-1,j} + (-1)^i \: \delta_{k+1,j}).
\end{align*}
\end{lemma}

\begin{proof}
Each follows from the definitions of $\ch_{k,i}$ and $\ch_0$, as well as the relationships:
\begin{align*}
	2 \cos k \theta \cos\theta &= \cos (k-1)\theta + \cos(k+1)\theta \\
	2 \sin k \theta \cos\theta &= \sin (k-1)\theta + \sin(k+1)\theta \\
	2 \cos k \theta \sin\theta &= - \sin (k-1)\theta + \sin(k+1)\theta \\
	2 \sin k \theta \sin\theta &= \cos (k-1)\theta - \cos(k+1)\theta.
\end{align*}
\end{proof}

\begin{lemma}\label{lemma:Qrecurrence} 
Define
\begin{align}
	\eta_{k} :=
		\begin{cases}
			0 &\text{if } k < 0 \\
			\ch_0 &\text{if } k = 0 \\
			\half &\text{otherwise}
		\end{cases}
\end{align}
$\scopnkia(x,y,z)$ satisfy the following recurrences:
\begin{align*}
	x \: \scopnkia(x,y,z) &= \alphaa_{n,k,1} \:  \scopa_{n-1, k-1, i}(x, y, z) + \alphaa_{n,k,2} \:  \scopa_{n-1, k+1, i}(x, y, z) \nonumber \\
		& \quad \quad + \alphaa_{n,k,3} \:  \scopa_{n, k-1, i}(x, y, z) + \alphaa_{n,k,4} \:  \scopa_{n, k+1, i}(x, y, z) \nonumber \\
		& \quad \quad + \alphaa_{n,k,5} \:  \scopa_{n+1, k-1, i}(x, y, z) + \alphaa_{n,k,6} \:  \scopa_{n+1, k+1, i}(x, y, z), \\ \\
	y \: \scopnkia(x,y,z) &= \betaa_{n,k,i,1} \:  \scopa_{n-1, k-1, |i-1|}(x, y, z) + \betaa_{n,k,i,2} \:  \scopa_{n-1, k+1, |i-1|}(x, y, z) \nonumber \\
		& \quad \quad + \betaa_{n,k,i,3} \:  \scopa_{n, k-1, |i-1|}(x, y, z) + \betaa_{n,k,i,4} \:  \scopa_{n, k+1, |i-1|}(x, y, z) \nonumber \\
		& \quad \quad + \betaa_{n,k,i,5} \:  \scopa_{n+1, k-1, |i-1|}(x, y, z) + \betaa_{n,k,i,6} \:  \scopa_{n+1, k+1, |i-1|}(x, y, z), \\ \\
	z \: \scopnkia(x,y,z) &= \gammaa_{n,k,1} \: \scopa_{n-1, k, i}(x, y, z) + \gammaa_{n,k,2} \: \scopa_{n, k, i}(x, y, z) + \gammaa_{n,k,3} \: \scopa_{n+1, k, i}(x, y, z),
\end{align*}
for $(x,y,z) \in \Omega$, where
\begin{align*}
	\alphaa_{n,k,1} &:= \eta_{k-1} \: \ip<\genjacnmk^{(a, 2k)}, \: \genjacnmk^{(a, 2(k-1))}>_{\genjacw^{(a, 2k)}}, \\
	\alphaa_{n,k,2} &:= \eta_{k} \: \ip<\genjacnmk^{(a, 2k)}, \genjac_{n-k-2}^{(a, 2(k+1))}>_{\genjacw^{(a, 2(k+1))}}, \\
	\alphaa_{n,k,3} &:=\eta_{k-1} \: \ip<\genjacnmk^{(a, 2k)}, \: \genjac_{n-k+1}^{(a, 2(k-1))}>_{\genjacw^{(a, 2k)}}, \\
	\alphaa_{n,k,4} &:= \eta_{k} \: \ip<\genjacnmk^{(a, 2k)}, \genjac_{n-k-1}^{(a, 2(k+1))}>_{\genjacw^{(a, 2(k+1))}}, \\
	\alphaa_{n,k,5} &:= \eta_{k-1} \: \ip<\genjacnmk^{(a, 2k)}, \: \genjac_{n-k+2}^{(a, 2(k-1))}>_{\genjacw^{(a, 2k)}}, \\
	\alphaa_{n,k,6} &:=\eta_{k} \: \ip<\genjacnmk^{(a, 2k)}, \genjacnmk^{(a, 2(k+1))}>_{\genjacw^{(a, 2(k+1))}}, \\
	\betaa_{n,k,i,j} &:= 
		\begin{cases}
			- \alphaa_{n,k,j} \quad &\text{if }(i = 0 \text{ and } j \text{ is odd}) \text{ or }(i = 1 \text{ and } j \text{ is even}) \\
			\alphaa_{n,k,j} \quad &\text{otherwise}
		\end{cases}, \\	
	\gammaa_{n,k,1} &:= \beta_{n-k-1}^{(a, 2k)}, \qquad \gammaa_{n,k,2} := \alpha_{n-k}^{(a, 2k)}, \qquad \gammaa_{n,k,3} := \beta_{n-k}^{(a, 2k)}.
\end{align*}
\end{lemma}
\begin{remark}
For $z$ multiplication, note that different Fourier modes do not interact. This is because $z$ is rotationally invariant. 
\end{remark}
\begin{proof}
The 3-term recurrence for multiplication by $z$ follows from (\ref{eqn:Rrecurrence}). For the recurrence for multiplication by $x$, since $\{\scopmjha\}$ for $m = 0,\dots,n+1$, $j = 0,\dots,m$, $h = 0,1$ is an orthogonal basis for any degree $n+1$ polynomial on $\Omega$, we can expand 
\begin{align*}
	x \: \scopnkia(x,y,z) = \sum_{m=0}^{n+1} \sum_{j=0}^m \sum_{h=0}^1 c_{m,j} \: \scopmjha(x,y,z).
\end{align*}
These coefficients are given by
\begin{align*}
	c_{m,j} = {\ip< x \: \scopnkia, \scopmjha >_{\scopa}}{\norm{\scopmjha}^{-2}_{\scopa}}
\end{align*}
where we show the non-zero coefficients that result are the $\alphaa_{n,k,1},\dots,\alphaa_{n,k,6}$ in the lemma.
Recall from equation (\ref{eqn:normscop}) that $\norm{\scopmjha}_{\scopa}^2 = \pi \: \normgenjac^{(a,2j)}$. Then for $m = 0,\dots,n+1$, $j = 0,\dots,m$, using a change of variables $(\cos\theta \sin\varphi, \: \sin\theta\sin\varphi, \: \cos\varphi) = (x, y, z)$:
\begin{align*}
	&\ip<x \: \scopnkia, \scopmjha>_{\scopa} \\
	&= \int_\Omega \scopnkia(\xvec,z) \: \scopmjha(\xvec,z) \: x \: \genjacw^{(a,0)}(z) \: \D A \\
	&= \Big( \int^1_\alpha \genjacnmk^{(a, 2k)}(z) \: \genjacmmj^{(a, 2j)}(z) \: \rho(z)^{k+j+1} \: \genjacw^{(a, 0)}(z) \: \D z \Big) \cdot \: \Big( \int_0^{2\pi} \chki(\theta) \: \chjh(\theta) \: \cos\theta \: \D \theta \Big) \\
	&= \Big( \int^1_\alpha \genjacnmk^{(a, 2k)}(z) \: \genjacmmj^{(a, 2j)}(z) \: \genjacw^{(a, k + j + 1)}(z) \: \D z \Big) \cdot \: \Big( \int_0^{2\pi} \chki(\theta) \: \chjh(\theta) \: \cos\theta \: \D \theta \Big) \\
	&= \half \: \pi \: \delta_{i,h} \: (\eta_{k-1} \: \delta_{k-1, j} + \eta_{k} \: \delta_{k+1, j})  \int^1_\alpha \genjacnmk^{(a, 2k)}(z) \: \genjacmmj^{(a, 2j)}(z) \: \genjacw^{(a, k + j + 1)}(z) \: \D z.
\end{align*}
where $\delta_{k, j}$ is the standard Kronecker delta function, using Lemma \ref{lemma:Yrecurrence}. Similarly, for the recurrence for multiplication by $y$, we can expand 
\begin{align*}
	y \: \scopnkia(x,y,z) = \sum_{m=0}^{n+1} \sum_{j=0}^m \sum_{h=0}^1 d_{m,j} \: \scopmjha(x,y,z).
\end{align*}
These coefficients are given by
\begin{align*}
	d_{m,j} = {\ip< y \: \scopnkia, \scopmjha >_{\scopa}}{\norm{\scopmjha}^{-2}_{\scopa}}
\end{align*}
where we show the non-zero coefficients that result are the $\betaa_{n,k,1},\dots,\betaa_{n,k,6}$ in the lemma:
\begin{align*}
	&\ip<y \: \scopnkia, \scopmjha>_{\scopa} \\
	&= \int_\Omega \scopnkia(\xvec,z) \: \scopmjha(\xvec,z) \: y \: \genjacw^{(a,0)}(z) \: \D A \\
	&= \Big( \int^1_\alpha \genjacnmk^{(a, 2k)}(z) \: \genjacmmj^{(a, 2j)}(z) \: \rho(z)^{k+j+1} \: \genjacw^{(a, 0)}(z) \: \D z \Big) \cdot \: \Big( \int_0^{2\pi} \chki(\theta) \: \chjh(\theta) \: \sin\theta \: \D \theta \Big) \\
	&= \Big( \int^1_\alpha \genjacnmk^{(a, 2k)}(z) \: \genjacmmj^{(a, 2j)}(z) \: \genjacw^{(a, k + j + 1)}(z) \: \D z \Big) \cdot \: \Big( \int_0^{2\pi} \chki(\theta) \: \chjh(\theta) \: \sin\theta \: \D \theta \Big) \\
	&= \half \: \pi \: \delta_{|i-1|,h} \: \big[(-1)^{i+1} \: \eta_{k-1} \: \delta_{k-1, j} + (-1)^i \: \eta_{k} \: \delta_{k+1, j}\big] \int^1_\alpha \genjacnmk^{(a, 2k)}(z) \: \genjacmmj^{(a, 2j)}(z) \: \genjacw^{(a, k + j + 1)}(z) \: \D z.
\end{align*}
where again $\delta_{k, j}$ is the standard Kronecker delta function, and we have used Lemma \ref{lemma:Yrecurrence}.

\end{proof}

The recurrences in Lemma \ref{lemma:Qrecurrence} lead to Jacobi operators that correspond to multiplication by $x$, $y$ and $z$. In later sections we will use an ordering of the OPs so that they are grouped by Fourier mode $k$, which is convenient for the application of differential and other operators to the vector of coefficients of a given function's expansion (some operators will exploit this ordering for operators where Fourier modes do not interact, and thus will be block-diagonal). Before that though, the ordering we will use in the remainder of this section is convenient for establishing Jacobi operators for multiplication by $x$, $y$ and $z$, and hence building the OPs and importantly obtaining the associated \textit{recurrence coefficient matrices} necessary for efficient function evaluation using the Clenshaw algorithm. In practice, it is simply a matter of converting coefficients between the two orderings. To this end, we define our OP-building ordering as follows. For $n=0,1,2,\dots$:
\begin{align*}
	\bigscoptna := 
		\begin{pmatrix}
			\scopa_{n,0,0}(x,y,z) \\
			\scopa_{n,1,0}(x,y,z) \\
			\scopa_{n,1,1}(x,y,z) \\
			\vdots \\
			\scopa_{n,n,0}(x,y,z) \\
			\scopa_{n,n,1}(x,y,z)
		\end{pmatrix} \in \R^{2n+1}, 
	\quad \quad 
	\bigscopta := 
		\begin{pmatrix}
			\bigscopta_0 \\
			\bigscopta_1 \\
			\bigscopta_2 \\
			\vdots \\
		\end{pmatrix}
\end{align*}
and set $ J_x^{(a)},  J_y^{(a)},  J_z^{(a)}$ as the Jacobi matrices corresponding to
\begin{align}
	J_x^{(a)} \: \bigscopta(x,y,z) = x \: \bigscopta(x,y,z), \nonumber \\
	J_y^{(a)} \: \bigscopta(x,y,z) = y \: \bigscopta(x,y,z), \label{eqn:jacobimatricesdefinitionalt} \\
	J_z^{(a)} \: \bigscopta(x,y,z) = z \: \bigscopta(x,y,z). \nonumber
\end{align}
where
\begin{align*}
	J_{x/y/z}^{(a)} &= 
		\begin{pmatrix}
			B^{(a)}_{x/y/z, 0} & A^{(a)}_{x/y/z, 0} & & & & \\
			C^{(a)}_{x/y/z, 1} & B^{(a)}_{x/y/z, 1} & A^{(a)}_{x/y/z, 1} & & & \\
			& C^{(a)}_{x/y/z, 2} & B^{(a)}_{x/y/z, 2} & A^{(a)}_{x/y/z, 2} & & & \\
			& & C^{(a)}_{x/y/z, 3} & \ddots & \ddots & \\
			& & & \ddots & \ddots & \ddots \\
		\end{pmatrix}.
\end{align*}
Note that $J_x^{(a)}, J_y^{(a)}, J_z^{(a)}$ are \textit{banded-block-banded matrices}:

\begin{definition}
	A block matrix $A$ with blocks $A_{i,j}$ has block-bandwidths $(L,U)$ if $A_{i,j} = 0$ for $- L \leq j-i \leq U$, and sub-block-bandwidths $(\lambda, \mu)$ if all blocks $A_{i,j}$ are banded with bandwidths $(\lambda,\mu)$. A matrix where the block-bandwidths and sub-block-bandwidths are small compared to the dimensions is referred to as a banded-block-banded matrix. 
\end{definition}

Each of these Jacobi matrices are then block-tridiagonal (block-bandwidths $(1,1)$). For $J_x^{(a)}$, the sub-blocks have sub-block-bandwidths $(2,2)$:
\begin{align*}
	A^{(a)}_{x,n} &:= 
		\begin{pmatrix}
			0 & A^{(a)}_{n,0,6} & 0 & & \\
			A^{(a)}_{n,1,5} & \ddots & \ddots & & \\
			& \ddots & \ddots & \ddots & \\
			& & A^{(a)}_{n,n,5} & 0 & A^{(a)}_{n,n,6} \\
		\end{pmatrix} \in \R^{(2n+1)\times(2n+3)}, \quad n = 0,1,2,\dots \\
	B^{(a)}_{x,n} &:= 
		\begin{pmatrix}
			0 & A^{(a)}_{n,0,4} & & \\
			A^{(a)}_{n,1,3} & \ddots & \ddots & \\
			& \ddots & \ddots & A^{(a)}_{n,n-1,4} \\
			& & A^{(a)}_{n,n,3} & 0
		\end{pmatrix} \in \R^{(2n+1)\times(2n+1)}  \quad n = 0,1,2,\dots \\
	C^{(a)}_{x,n} &:= 
		\begin{pmatrix}
			0 & A^{(a)}_{n,0,2} & & \\
			A^{(a)}_{n,1,1} & \ddots & \ddots & \\
			& \ddots & \ddots &A^{(a)}_{n,n-2,2} \\
			& & \ddots & 0 \\
			& & & A^{(a)}_{n,n,1} \\
		\end{pmatrix} \in \R^{(2n+1)\times(2n-1)}, \quad n = 1,2,\dots
\end{align*}
where for $k = 1,\dots,N, \: n = k,\dots,N$
\begin{align*}
	A^{(a)}_{n,k,j} &:= 
		\begin{pmatrix}
			\alphaa_{n,k,j} & 0 \\
			0 & \alphaa_{n,k,j}
		\end{pmatrix} \in \R^{2\times2}, (k \ne 1 \text { for } j \text{ odd}) \\
	A^{(a)}_{n,0,j} &:=
		\begin{pmatrix}
			\alphaa_{n,0,j} & 0
		\end{pmatrix} \in \R^{1\times2}, j \text{ even} \\
	A^{(a)}_{n,1,j} &:=
		\begin{pmatrix}
			\alphaa_{n,1,j} \\
			0
		\end{pmatrix} \in \R^{2\times1}, j \text{ odd}.
\end{align*}
For $J_y^{(a)}$, the sub-blocks have sub-block-bandwidths $(3,3)$:
\begin{align*}
	A^{(a)}_{y,n} &:= 
		\begin{pmatrix}
			0 & B^{(a)}_{n,0,6} & 0 & & \\
			B^{(a)}_{n,1,5} & \ddots & \ddots & & \\
			& \ddots & \ddots & \ddots & \\
			& & B^{(a)}_{n,n,5} & 0 & B^{(a)}_{n,n,6} \\
		\end{pmatrix} \in \R^{(2n+1)\times(2n+3)}, \quad n = 0,1,2,\dots \\
	B^{(a)}_{y,n} &:= 
		\begin{pmatrix}
			0 & B^{(a)}_{n,0,4} & & \\
			B^{(a)}_{n,1,3} & \ddots & \ddots & \\
			& \ddots & \ddots & B^{(a)}_{n,n-1,4} \\
			& & B^{(a)}_{n,n,3} & 0
		\end{pmatrix} \in \R^{(2n+1)\times(2n+1)}  \quad n = 0,1,2,\dots \\
	C^{(a)}_{y,n} &:= 
		\begin{pmatrix}
			0 & B^{(a)}_{n,0,2} & & \\
			B^{(a)}_{n,1,1} & \ddots & \ddots & \\
			& \ddots & \ddots & B^{(a)}_{n,n-2,2} \\
			& & \ddots & 0 \\
			& & & B^{(a)}_{n,n,1} \\
		\end{pmatrix} \in \R^{(2n+1)\times(2n-1)}, \quad n = 1,2,\dots
\end{align*}
where for $k = 1,\dots,N, \: n = k,\dots,N$  
\begin{align*}
	B^{(a)}_{n,k,j} &:= 
		\begin{pmatrix}
			0 & \betaa_{n,k,0,j} \\
			\betaa_{n,k,1,j} & 0
		\end{pmatrix} \in \R^{2\times2}, (k \ne 1 \text { for } j \text{ odd}) \\
	B^{(a)}_{n,0,j} &:=
		\begin{pmatrix}
			0 & \betaa_{n,0,0,j}
		\end{pmatrix} \in \R^{1\times2}, j \text{ even} \\
	B^{(a)}_{n,1,j} &:=
		\begin{pmatrix}
			0 \\
			\betaa_{n,1,1,j}

>>>>

\title{Sparse spectral methods for partial differential equations on spherical caps, highlights}
\author{Ben Snowball, Sheehan Olver\thanks{Department of Mathematics, Imperial College, London, England}}
%\date{}							% Activate to display a given date or no date

@@

\begin{enumerate}
\item Constructs sparse discretisations for partial differential operators on spherical caps using orthogonal polynomials. 
\item Approximations are observed to converge spectrally fast.
\item Optimal complexity is achieved for operators that are invariant to rotation around the pole.
\item Examples include (surface) Poisson equation, Helmholtz equation with variable coefficients, and the biharmonic equation.
\end{enumerate}
